Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-01-31 WholePlotTrajectories.Rmd
intro

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the resilience of community taxonomic and functional diversity and composition remains controversial. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed tree community trajectories following a disturbance gradient in a Neotropical forest. We examined community diversity, composition and redundancy trajectories over 30 years, considering both taxonomic and functional trajectories based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::word_document2: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
---

@@

The IDH then assumes both predictable trajectories depending on the disturbance intensity, and, without additional disturbance, the recovery of pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Sheil2003;@Willig2018].
In highly diverse tropical forests, however, the validation of the IDH remains controversial [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hugues2007].
Empirical evidence generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which might give limited or misleading information on community disturbance response. 
The IDH pattern of diversity peak under an intermediate disturbance regime can indeed result from various ecological processes [@Sheil2003].
To disentangle these various processes and grasp the different facets of community response to disturbance it is crucial to combine a suitable range of approaches [@Shea2004].
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) examined community recovery time.

@Rees2001].
Yet, observed trajectories proved to often deviate from predictions and from the recovery of initial state and thus evidenced the interplay of stochastic processes [@Hubbel2001;@Norden2015; @Norden2017].
The debate underlying the study of community response to disturbance is now on the relative importance of stochastic and deterministic processes, and on the nature of ecological rules and mechanisms they generate.

Specifically, the determinism of post-disturbance trajectories is the base of the intermediate disturbance hypothesis (IDH) which is assumed to be a determinant of tropical forest biodiversity [@Connell1978].

@@

Our study highlighted the combination of deterministic and stochastic processes determining the succession of recruited species and shaping community response to disturbance. 
The deterministic recruitment of pioneers shaped functional trajectories convergent in the functional space, while dispersal limitations shaped stochastic taxonomic trajectories maintaining community divergence in taxonomic composition.
In accordance with the IDH, community taxonomic diversity was maximise for intermediate disturbance intensity, before reaching an intensity threshold beyond which the taxonomic richness decreased and the taxonomic evenness remained persistently higher.
Conversely, functional trajectories proved decoupled from the taxonomic ones and did not follow the hypothesis of the IDH as no distinct trajectories emerged with increasing disturbance intensity. 
The decoupling between taxonomic and functional disturbance was explained by a decrease of community functional redundancy mitigating the functoinal impact of disturbance.
The resilience of tropical forests, defined in terms of recovery to pre-disturbance state, proved tangible but requiring several decades. 
Still, the disturbance impact on community redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities  [@Herault2018].

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Bruno Herault
    authsuperscript: 2
  - name: Eric Marcon
    authsuperscript: 1
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: INPHB, Institut National Polytechnique Félix Houphoüet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: https://github.com/ArianeMirabel
abstract: 
 Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
 The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the resilience of community taxonomic and functional diversity and composition remains controversial. 
 To disentangle the ecological processes driving community response to disturbance, we analyzed tree community trajectories following a disturbance gradient in a Neotropical forest. We examined community diversity, composition and redundancy trajectories over 30 years, considering both taxonomic and functional trajectories based on 7 leaf, stem and life-history traits. We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities. The IDH did predict community taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity. 
 Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.
keywords: [Community Ecology, Community Diversity Determinants, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Mid-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity]
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings re-or-gan-i-za-tion post-dis-tur-bance dis-tur-bance}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
  bookdown::gitbook: default
---

@@

The IDH then assumes both predictable trajectories depending on the disturbance intensity, and, without additional disturbance, the recovery of pre-disturbance state defined by the dominance of late-successional species [@Rees2001;@Sheil2003;@Willig2018].
In highly diverse tropical forests, however, the validation of the IDH remains controversial [@Fox2013;@Sheil2013] and observed post-disturbance trajectories are often blurry compared to theoretical expectations [@Hubbel2001;@Hugues2007;@Norden2015;@Norden2017].
Empirical evidence generally rely on the analysis of a single diversity metric, often the taxonomic richness [@Molino2001], which might give limited or misleading information on community disturbance response. 
The IDH pattern of diversity peak under an intermediate disturbance regime can indeed result from various ecological processes [@Sheil2003].
To disentangle these various processes and grasp the different facets of community response to disturbance it is crucial to combine a suitable range of approaches [@Shea2004].
More ecological-meaningful analysis would first couple taxonomic richness and evenness, which reveals the changes in species abundance distribution and hence in the underlying ecological processes [@Chaudhary2016]. 
Analysis would as well include taxonomic composition, which is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. 
Eventually, including a functional approach accounting for species biological attributes would be highly revealing of ecological process involved in community response to disturbance. The functional trait-based approach indeed directly links community diversity, composition and redundancy to ecosystem functioning and environmental constraints [@Violle2007b;@Baraloto2012a]. 
In that respect, a vast literature allowed recognizing major traits related to species ecology and mediating species performance in a given environment [@Diaz2005]. 
Specifically in tropical forests, the functional approach revealed the post-disturbance deterministic processes entailing a shift from a dominance of “conservative” slow-growing species dealing with scarce resources, to a dominance of “acquisitive” fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011]. 
This shift is translated into the trajectories of average community value of key functional traits related to resource acquisition (leaf and stem traits) and life-history strategy (seed mass, maximum size)  [@Wright2004;@Diaz2005;@TerSteege2006;@Westoby2006a;@Chave2009b]. 
The functional approach would eventually be completed with the analysis of functional redundancy that quantifies the amount of shared trait values among species [@Carmona2016]. The high functional redundancy of hyper-diverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of Neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. Specifically, we elucidateed (i) the community taxonomic and functional recovery and (ii) the ecological processes underlying community post-disturbance trajectories and discussed the validity and the extent of the IDH in the long term for tropical forest, and (iii) examined community recovery time.

@@

Our study highlighted the combination of deterministic and stochastic processes determining the succession of recruited species and shaping community response to disturbance. 
The deterministic recruitment of pioneers shaped functional trajectories convergent in the functional space, while dispersal limitations shaped stochastic taxonomic trajectories maintaining community divergence in taxonomic composition.
In accordance with the IDH, community taxonomic diversity was maximised for intermediate disturbance intensity, before reaching an intensity threshold beyond which the taxonomic richness decreased and the taxonomic evenness remained persistently higher.
Conversely, functional trajectories proved decoupled from the taxonomic ones and did not follow the hypothesis of the IDH as no distinct trajectories emerged with increasing disturbance intensity. 
The decoupling between taxonomic and functional disturbance was explained by a decrease of community functional redundancy mitigating the functoinal impact of disturbance.
The resilience of tropical forests, defined in terms of recovery to pre-disturbance state, proved tangible but requiring several decades. 
Still, the disturbance impact on community redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Herault2018].

### 2019-01-31 WholePlotTrajectories.tex
intro

The IDH then assumes both predictable trajectories depending on the
disturbance intensity, and, without additional disturbance, the recovery
of pre-disturbance state defined by the dominance of late-successional
species \citep{Rees2001, Sheil2003, Willig2018}. In highly diverse
tropical forests, however, the validation of the IDH remains
controversial \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hugues2007}. Empirical evidence generally rely on the analysis of
a single diversity metric, often the taxonomic richness
\citep{Molino2001}, which might give limited or misleading information
on community disturbance response. The IDH pattern of diversity peak
under an intermediate disturbance regime can indeed result from various
ecological processes \citep{Sheil2003}. To disentangle these various
processes and grasp the different facets of community response to
disturbance it is crucial to combine a suitable range of approaches
\citep{Shea2004}. More ecological-meaningful analysis would first couple
taxonomic richness and evenness, which reveals the changes in species
abundance distribution and hence in the underlying ecological processes
\citep{Chaudhary2016}. Analysis would as well include taxonomic
composition, which is crucial for conservation issues
\citep{Lavorel2002, Bellwood2006}. Eventually, including a functional
approach accounting for species biological attributes would be highly
revealing of ecological process involved in community response to
disturbance. The functional trait-based approach indeed directly links
community diversity, composition and redundancy to ecosystem functioning
and environmental constraints \citep{Violle2007b, Baraloto2012a}. In
that respect, a vast literature allowed recognizing major traits related
to species ecology and mediating species performance in a given
environment \citep{Diaz2005}. Specifically in tropical forests, the
functional approach revealed the post-disturbance deterministic
processes entailing a shift from a dominance of ``conservative''
slow-growing species dealing with scarce resources, to a dominance of
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of average community value of key
functional traits related to resource acquisition (leaf and stem traits)
and life-history strategy (seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

@@

\citet{Rees2001}{]}. Yet, observed trajectories proved to often deviate
from predictions and from the recovery of initial state and thus
evidenced the interplay of stochastic processes
\citep{Hubbel2001, Norden2015, Norden2017}. The debate underlying the
study of community response to disturbance is now on the relative
importance of stochastic and deterministic processes, and on the nature
of ecological rules and mechanisms they generate.

Specifically, the determinism of post-disturbance trajectories is the
base of the intermediate disturbance hypothesis (IDH) which is assumed
to be a determinant of tropical forest biodiversity \citep{Connell1978}.

@@

Our study highlighted the combination of deterministic and stochastic
processes determining the succession of recruited species and shaping
community response to disturbance. The deterministic recruitment of
pioneers shaped functional trajectories convergent in the functional
space, while dispersal limitations shaped stochastic taxonomic
trajectories maintaining community divergence in taxonomic composition.
In accordance with the IDH, community taxonomic diversity was maximise
for intermediate disturbance intensity, before reaching an intensity
threshold beyond which the taxonomic richness decreased and the
taxonomic evenness remained persistently higher. Conversely, functional
trajectories proved decoupled from the taxonomic ones and did not follow
the hypothesis of the IDH as no distinct trajectories emerged with
increasing disturbance intensity. The decoupling between taxonomic and
functional disturbance was explained by a decrease of community
functional redundancy mitigating the functoinal impact of disturbance.
The resilience of tropical forests, defined in terms of recovery to
pre-disturbance state, proved tangible but requiring several decades.
Still, the disturbance impact on community redundancy cautioned against
the risks of infrequent species loss and the persistence of
disturbance-specific communities \citep{Herault2018}.

>>>>

The IDH then assumes both predictable trajectories depending on the
disturbance intensity, and, without additional disturbance, the recovery
of pre-disturbance state defined by the dominance of late-successional
species \citep{Rees2001, Sheil2003, Willig2018}. In highly diverse
tropical forests, however, the validation of the IDH remains
controversial \citep{Fox2013, Sheil2013} and observed post-disturbance
trajectories are often blurry compared to theoretical expectations
\citep{Hubbel2001, Hugues2007, Norden2015, Norden2017}. Empirical
evidence generally rely on the analysis of a single diversity metric,
often the taxonomic richness \citep{Molino2001}, which might give
limited or misleading information on community disturbance response. The
IDH pattern of diversity peak under an intermediate disturbance regime
can indeed result from various ecological processes \citep{Sheil2003}.
To disentangle these various processes and grasp the different facets of
community response to disturbance it is crucial to combine a suitable
range of approaches \citep{Shea2004}. More ecological-meaningful
analysis would first couple taxonomic richness and evenness, which
reveals the changes in species abundance distribution and hence in the
underlying ecological processes \citep{Chaudhary2016}. Analysis would as
well include taxonomic composition, which is crucial for conservation
issues \citep{Lavorel2002, Bellwood2006}. Eventually, including a
functional approach accounting for species biological attributes would
be highly revealing of ecological process involved in community response
to disturbance. The functional trait-based approach indeed directly
links community diversity, composition and redundancy to ecosystem
functioning and environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, a vast literature
allowed recognizing major traits related to species ecology and
mediating species performance in a given environment \citep{Diaz2005}.
Specifically in tropical forests, the functional approach revealed the
post-disturbance deterministic processes entailing a shift from a
dominance of ``conservative'' slow-growing species dealing with scarce
resources, to a dominance of ``acquisitive'' fast-growing species with
rapid and efficient use of abundant resources
\citep{Reich2014, Herault2011}. This shift is translated into the
trajectories of average community value of key functional traits related
to resource acquisition (leaf and stem traits) and life-history strategy
(seed mass, maximum size)
\citep{Wright2004, Diaz2005, TerSteege2006, Westoby2006a, Chave2009b}.
The functional approach would eventually be completed with the analysis
of functional redundancy that quantifies the amount of shared trait
values among species \citep{Carmona2016}. The high functional redundancy
of hyper-diverse tropical forests \citep{Bellwood2006} mitigates the
impacts of species removal on ecosystem functioning and determines the
resilience of communities after disturbance
\citep{Elmqvist2003, Diaz2005}.

@@

@@

Our study highlighted the combination of deterministic and stochastic
processes determining the succession of recruited species and shaping
community response to disturbance. The deterministic recruitment of
pioneers shaped functional trajectories convergent in the functional
space, while dispersal limitations shaped stochastic taxonomic
trajectories maintaining community divergence in taxonomic composition.
In accordance with the IDH, community taxonomic diversity was maximised
for intermediate disturbance intensity, before reaching an intensity
threshold beyond which the taxonomic richness decreased and the
taxonomic evenness remained persistently higher. Conversely, functional
trajectories proved decoupled from the taxonomic ones and did not follow
the hypothesis of the IDH as no distinct trajectories emerged with
increasing disturbance intensity. The decoupling between taxonomic and
functional disturbance was explained by a decrease of community
functional redundancy mitigating the functoinal impact of disturbance.
The resilience of tropical forests, defined in terms of recovery to
pre-disturbance state, proved tangible but requiring several decades.
Still, the disturbance impact on community redundancy cautioned against
the risks of infrequent species loss and the persistence of
disturbance-specific communities \citep{Herault2018}.
